---
   author: "Yves Deville"
   date: "`r Sys.Date()`"
   title: "Special Values in Extreme-Value Distributions" 
   output: 
      html_document:
        toc: true
---

# Probability functions of Extreme-Value distributions

Many R packages define the probability functions (density,
distribution and quantile) of the Generalised Pareto (GP) and the
Generalised Extreme Value distribution (GEV). While some packages use
these functions internally, most of them also export them as well. As
a result, the R user interested by using such a probability function
has to choose between the different implementations. The present
vignette inquires about quite subtle details in the way the functions
work. In most cases, these details can not be anticipated by reading
the package documentation.

## Vectorisation

While in most (if not all) of the packages the first argument of the
probability function is vectorised, this is generally not the case for
the other arguments representing the parameters (location, scale,
shape). This constrasts with the classical probability functions
implemented in the **stats** package, in which the parameter arguments
are vectorised.

```{r}
dnorm(1:4, mean = 1:4)
dnorm(1:4, sd = 1:4)
```

```{r}
try(evd::dgev(x = c(1.4, 1.5), shape = c(-0.1, -0.1)))
try(SpatialExtremes::dgev(x = c(1.4, 1.5), shape = c(-0.1, -0.1)))
try(mev::dgev(x = c(1.2, 1.3), shape = c(0.2, -0.1)))
```

A vectorisation w.r.t. the parameter arguments is a plus at least in
the Bayesian framework when the probability functions have to be
evaluated using vectors of MCMC iterates for the parameters. Using
loops coded in R language may then lead to slowly executing code.



## Non-finite values 

By non-finite values we mean the four special values for a `double`
`-Inf`, ` Inf`, `NA` and `NaN`. Note that `NA` is used to show a
missing value whatever be its type, but there are different classes of
missing values. By default `NA` refers to the logical missing value
while we will here be concerned by numeric missing values.

```{r}
x <- NA_real_
class(x)
y <- NA
class(y)
class(Inf)
class(-Inf)
class(NaN)
``` 

### Non-finite values in the first argument

Most probability functions from EV packages accept non-finite values
in their first arguments, with a few exceptions
```{r, POT-dgpd-error}
try(POT::dgpd(x = c(0, Inf, NA), scale = 1))
```
### Non-finite values in the parameter arguments

As for non-finite values for the parameter arguments, only a few
packages allow this possibility in a consistent way. Beside **nieve**,
the **texmex** package seems to be the sole package both for the
density and the distribution functions.

For some uses, this feature can be important because the density and the
distributions functions are often used to define a log-likelihood
function that has to be maximised numerically. By experience we know
that the optimisation may have to evaluate the objective at a vector
of parameters embeding non-finite values, without compromising an
eventual convergence. Moreover, it is a desirable feature of an
optimisation routine to allow the evaluation of the objective or
constraint at such vector.  The `stats::optim` routine allows this
when used with `method = "BFGS"`, ant this is a very popular
choice. Optimisation routines from the **nloptr** package allow the
parameter to take non-finite values both in the objective and in the
constraint function, if any. Of course the log-likelihood could catch
non-finite values before calling the probability function(s), but this
is an extra pain and may slow down the estimation

### "Classical" behaviour for non-finite values

Remind of how the probability function of the classical distribution
in the **stats** package work.

- The first argument of the density, distribution of quantile function
  is a vector in which some elements can be non-finite without casting
  any error or warning. In the corresponding returned vector, an input
  value `NA` (resp. `NaN`) will produce an output value `NA`
  (resp. `NaN`). An infinite value `-Inf` (resp. `Inf`) correspond to:
  an output value `0` (resp. `0`) for the density function, an output
  value `0` (resp. `1`) for the distribution function and `NaN`
  (resp. `NaN`) for the quantile function.
  
- When the first argument `p` of the quantile function contains a
  value `< 0.` or ` > 1.` a warning is cast and the corresponding
  output value is `NaN`. When `p` is `0.` (resp.  `1.`), the
  corresponding output value is the lower (resp. upper) end-point
  of the distribution be it finite or not.
  
- The parameter arguments can be vectors and some elements in these be
  non-finite without resulting in an error. An input value `NA`
  (resp. `NaN`) will produce an output value `NA` (resp. `NaN`).  When
  an input value is `-Inf` or `Inf` the corresponding value will
  usually be `NaN`. Yet it can be a finite value in some cases
  as in `dnorm(1.3, mean = Inf)` which returns `r  dnorm(1.3, mean = Inf)`.

- When incorrect values are given in a parameter argument such as 
  a negative value for a scale argument, the corresponding output value 
  is `NaN`.

# Extreme-Value packages

For a number of Extreme-Values packages we evaluate the results given
by the density function for their two-parameter GP and for their
three-parameter GEV distributions if any. Note that each row in a
table corresponds to a call using vectorised w.r.t. thte first
argument (`x` for a density, `q` for a distribution and `p` for a
quantile). So an error must be considered as holding for a full row.

## Findings

Our findings are as follows.

- For all packages the quantile function `q` casts an error when the
  vector given as probability argument contains non-finite
  values. This behaviour differs from that of the quantile
  function for classical probability functions.

- In the packages **evd**, **`POT`** and **SpatialExtremes**, the GPD
  density unduly casts an error when the shape is `0.0` and `x` is a
  vector containing non-finite values. This may be considered as a
  bug.

- Most packages do not cope with non-finite values for the parameters
  as the classical distribution functions do. For some packages the
  behaviour is not consistent between the two distributions or between
  the density and the distribution functions for a same distribution.


```{r init, echo=FALSE} 
library(nieve)
library(data.table)

Pars <- list("GEV" = c("loc", "scale", "shape"),
             "GPD2"= c("scale", "shape"))

ProbFuns <- list("d" = list(name = "density", arg = "x"),
                 "p" = list(name = "distribution", arg = "q"),
                 "q" = list(name = "quantile", arg = "p"))

ArgValues <- list("GEV" = list(d = c(1.4, -Inf, Inf, NA, NaN),
                               p = c(1.5, -Inf, Inf, NA, NaN),
                               q = c(-0.1, 0.0, 0.5, 1.0, 1.1, -Inf, Inf, NA, NaN),
                               loc = c(0.0, - Inf, Inf, NA, NaN)),
                  "GPD2" = list(d = c(1.4, -Inf, Inf, NA, NaN),
                                p = c(1.5, -Inf, Inf, NA, NaN),
                                q = c(-0.1, 0.0, 0.5, 1.0, 1.1, -Inf, Inf, NA, NaN)))
## Describe the values of the parameters to be used.
ParValues <- list("GEV" = list(loc = c(0.0, - Inf, Inf, NA, NaN),
                               scale = c(1.0, -1.0, 0.0, - Inf, Inf, NA, NaN),
                               shape = c(-0.4, 0.0, 0.5, - Inf, Inf, NA, NaN)),
                  "GPD2" = list(scale = c(1.0, -1.0, 0.0, - Inf, Inf, NA, NaN),
                                shape = c(-0.4, 0.0, 0.5, - Inf, Inf, NA, NaN)))

Altern <- list("nieve-GEV" = list(nieve = "GEV",
                                  Suff = "GEV",
                                  Pnm = c(loc = "loc",
                                          scale = "scale",
                                          shape = "shape")),
               "nieve-GPD2" = list(nieve = "GPD2",
                                   Suff = "GPD2",
                                   Pnm = c(scale = "scale",
                                           shape = "shape")),
               "mev-GEV" = list(nieve = "GEV",
                                Suff = "gev",
                                Pnm = c(loc = "loc",
                                        scale = "scale",
                                        shape = "shape")),
               "mev-GPD2" = list(nieve = "GPD2",
                                 Suff = "gp",
                                 Pnm = c(scale = "scale",
                                         shape = "shape")),
               "evd-GEV" = list(nieve = "GEV",
                                Suff = "gev",
                                Pnm = c(loc = "loc",
                                        scale = "scale",
                                        shape = "shape")),
               "evd-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale",
                                        shape = "shape")),
               "POT-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale", shape = "shape")),
               "SpatialExtremes-GPD2" = list(nieve = "GPD2",
                                             Suff = "gpd",
                                             Pnm = c(scale = "scale",
                                                     shape = "shape")),
               "SpatialExtremes-GEV" = list(nieve = "GEV",
                                           Suff = "gev",
                                           Pnm = c(loc = "loc",
                                                   scale = "scale",
                                                   shape = "shape")),
               "tea-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale",
                                         shape = "shape")),
               "texmex-GEV" = list(nieve = "GEV",
                                 Suff = "gev",
                                 Pnm = c(loc = "mu",
                                         scale = "sigma",
                                         shape = "xi"),
                                 Def = c(mu = 0.0, sigma = 1.0, xi = 0.0)),
               "texmex-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "sigma", shape = "xi"),
                                 Def = c(sigma = 1.0, xi = 0.0)),
               "extRemes-GPD2" = list(nieve = "GPD2",
                                      Suff = "evd",
                                      Pnm = c(scale = "scale", shape = "shape"),
                                      Extra = list(type = "GP")),
               "extRemes-GEV" = list(nieve = "GEV",
                                     Suff = "evd",
                                     Pnm = c(loc = "loc",
                                             scale = "scale",
                                             shape = "shape"),
                                     Extra = list(type = "GEV")),
               "revdbayes-GPD2" = list(nieve = "GPD2",
                                       Suff = "gp",
                                       Pnm = c(scale = "scale", shape = "shape")),
               "revdbayes-GEV" = list(nieve = "GEV",
                                      Suff = "gev",
                                      Pnm = c(loc = "loc",
                                             scale = "scale",
                                             shape = "shape"))
)

A <- list()
for (i in seq_along(Altern)) {
    A[[i]] <- data.frame(nieve = Altern[[i]]$nieve,
                         Pkg = gsub("-.*$", "", names(Altern)[i]),
                         Suff = Altern[[i]]$Suff,
                         Pnm = Altern[[i]]$Pnm)
}
AlternDf <- rbindlist(A)
AlternDf <- within(AlternDf, Pkg <- factor(Pkg))

testNonFinite <- function(pkg, print = 0) {

    LMat <- list()
    
    if (print) {
        cat("\n======================\n")
        cat("Package: ", pkg, "\n")
        cat("======================\n")
    }

    dfi <- subset(AlternDf, Pkg == pkg)
    dfi <- dfi[order(dfi$nieve), ]
    
    for (distName in unique(dfi$nieve)) {
        ind <- (dfi$nieve == distName)
        if (print) {
            cat("\no Dist: ", distName, "\n")
        }
        pns <- Pars[[distName]]
        for (pn in pns) {
            if (print) {
                cat("\n  - Parameter:   ", pn, "\n")
            }
            pv <- ParValues[[distName]][[pn]]
            if (print) {
                cat("    Par. values:   ", pv, "\n")
            }
            for (pf in c("d", "p", "q")) {
                ev <- ArgValues[[distName]][[pf]]
                if (print) {
                    cat("\n      Probability function:   ", pf, "\n")
                }
                ## cat("         First argument:   ", format(ev), "\n")
                funAlt <- get(paste0(pf, dfi[ind, ]$Suff[1]),
                              mode = "function",
                              envir = getNamespace(pkg))
                ## cat("         Results:   \n")
                mat <- matrix("", nrow = length(pv), ncol = length(ev))
                rownames(mat) <- paste("       ", pn, "=", format(pv, trim = TRUE))
                colnames(mat) <- paste(ProbFuns[[pf]]$arg, "=", format(ev, trim = TRUE))
                
                for (j in seq_along(pv)) {
                    
                    ## Do we need an ersatz for default values?
                    LI <- Altern[[paste(pkg, distName, sep = "-")]]
                    if (!is.null(LI$Def)) {
                        argAlt <- as.list(LI$Def)
                        argAlt[LI$Pnm[pn]] <- pv[j]
                        add <- list(x = ev)
                        names(add) <- ProbFuns[[pf]]$arg
                        argAlt <- c(add, argAlt)
                    } else {
                        argAlt <- list(ev, pv[j])
                        names(argAlt) <- c(ProbFuns[[pf]]$arg, pn)
                    }
                    
                    if (!is.null(LI$Extra)) {
                       argAlt <- c(argAlt, LI$Extra)
                    }
                    
                    resAlt <- try(do.call(funAlt, argAlt), silent = TRUE)
                    if (!inherits(resAlt, "try-error")) {
                        res <- paste(format(resAlt))
                        mat[j, ] <- format(resAlt)
                    } else {
                        res <- "ERROR"
                        mat[j, ] <- "*****"
                        mat[j, 1] <- "ERROR"
                    }
                }
                LMat[[paste0(pf, distName)]] <- noquote(mat)
                attr(LMat[[paste0(pf, distName)]], "name") <-
                    paste0(pkg, "::", pf, dfi[ind, ]$Suff[1])
            }
        }
    }
    LMat
}
```


```{r, echo=FALSE, results='asis'}
src <- lapply(levels(AlternDf$Pkg), function(pkg) {
    knitr::knit_expand('templates/nonFiniteResults.Rmd')
})
res <- knitr::knit_child(text = unlist(src), quiet = TRUE)
cat(res, sep = '\n')
```
