---
   author: "Yves Deville"
   date: "`r Sys.Date()`"
   title: "Special Values in Extreme-Value Distributions" 
   output: 
      pdf_document:
        toc: true
---

```{r, echo=FALSE, results="hide"} 
Sys.setenv(LANGUAGE = "en")
```

# Note 

This vignette has been compiled using **nieve** `r packageVersion("nieve")` 
and `r R.Version()$version.string`. The version of the packages used
is mentioned with the results. This is a preliminary version and a
*References* section duly citing the packages used will be added soon.

# Probability functions of Extreme-Value (EV) distributions

Many R packages define the probability functions (density,
distribution and quantile) of the Generalized Pareto (GP) and the
Generalized Extreme Value distribution (GEV). While these packages
could have kept the probability functions as for internal use, they
often instead export the functions. As a result, the R user interested
by using such a probability function has to choose between the
different implementations. The present vignette inquires about quite
subtle details in the way the functions work. In most cases, these
details can not be anticipated by reading the package documentation.

In this vignette we focus on the two-parameter GP and the three
parameter GEV distributions. These will be denoted as `GPD2` and `GEV`
as in **nieve**, although the name of these vary across packages. In
all packages the three functions of interest for each of the
distribution have their name obtained by pasting a prefix to the
distribution name. These are

- The density function with prefix `"d"` and its first argument 
  named `x`.
- The distribution function with prefix `"p"` and with its first 
  argument named `q`.
- The quantile function with prefix `"q"` and with its first argument
  named `p`.

The other arguments are the parameters corresponding to the location
(GEV) and the scale and the shape. These arguments are often named
`loc`, `scale` and `shape`, but not always. Also, default values for
the parameter are most often provided as `loc = 0.0`, `scale = 1.0`,
and `shape = 0.0`, but there are some exceptions. We may refer to
these values as the "standard" default values.

Note that **evd** was first released on CRAN in 2002, and has been a
source of inspiration for later packages, in which the probability
functions can be nearly copies of those of **evd**. The package
**ismev** is also one of the first released on CRAN. While some of the
older packages have been re-factored (**extRemes**) only limited
changes were made in the code of the package **evd**.


## Vectorization

While in most (if not all) of the packages the first argument of the
probability function is vectorized, this is generally not the case for
the other arguments representing the parameters (location, scale,
shape). This contrasts with the classical probability functions
implemented in the **stats** package, in which the parameter arguments
are vectorized.

```{r}
dnorm(1:4, mean = 1:4)
dnorm(1:4, sd = 1:4)
```

```{r}
try(evd::dgev(x = c(1.4, 1.5), shape = c(0.1, -0.1)))
try(SpatialExtremes::dgev(x = c(1.4, 1.5), shape = c(0.1, -0.1)))
try(mev::dgev(x = c(1.2, 1.3), shape = c(0.2, -0.1)))
```

A vectorization w.r.t. the parameter arguments is a plus at least in
the Bayesian framework when the probability functions have to be
evaluated using vectors of MCMC iterates for the parameters. Using
loops coded in R language may then lead to slowly executing code.  In
the **revdbayes** package the two distributions GPD2 and GEV are
suitably vectorized w.r.t. the parameters.

## Non-finite values 

By non-finite values we mean the four special values for a `double`
`-Inf`, ` Inf`, `NA` and `NaN`. Note that `NA` is used to show a
missing value whatever be its type, but there are different classes of
missing values. By default `NA` refers to the logical missing value
while we will here be concerned by numeric missing values.

```{r}
x <- NA_real_
class(x)
y <- NA
class(y)
class(Inf)
class(-Inf)
class(NaN)
``` 

### Non-finite values in the first argument

Most probability functions from EV packages accept non-finite values
in their first arguments, with a few exceptions or problems

```{r, POT-dgpd-error, message=FALSE, warning=FALSE}
try(evd::dgpd(c(0.0, Inf)))
try(evd::dgpd(c(0.0, 1.0, Inf)))

```

### Non-finite values in the parameter arguments

As for non-finite values of the parameter arguments, only a few
packages allow this possibility in a consistent way. Beside **nieve**,
the **texmex** package seems to be the sole package conforming to the
classical behaviour as described below both for the density and the
distribution functions.

Accepting non-finite values of the parameter arguments can be
important because the density and the distribution functions are often
used to define a log-likelihood function that has to be maximised
numerically. By experience we know that the optimisation may have to
evaluate the objective at a vector of parameters embedding non-finite
values, without compromising an eventual convergence. Moreover, it is
a desirable feature of an optimisation routine to allow the evaluation
of the objective or constraint at such vector.  The `stats::optim`
routine allows this when used with `method = "BFGS"`, and this is a
very popular choice. Optimisation routines from the **nloptr** package
allow the parameter to take non-finite values both in the objective
and in the constraint function, if any. Of course the log-likelihood
function written by the user could catch non-finite values before
calling the probability function(s), but this is an extra pain and may
slow down the estimation. More importantly, it seems a good practice
that the log-likelihood function of a "fitted model" object remains
consistent with the exported distribution functions. For instance the
log-likelihood of a model specifying i.i.d. observations should be
equal to the sum of the log-density values at the observations as
computed from the exported density function. In other words the
function implementing the log-likelihood (be it coded in R or in a
compiled language) should call the proability functions as exported by
the package.

### "Classical" behaviour for non-finite values

Remind of how the probability functions of the classical continuous
distributions in the **stats** package work.

1. The first argument of the density, distribution of quantile
   function is a vector in which some elements can be non-finite
   without casting any error or warning. In the corresponding returned
   vector, an input value `NA` (resp. `NaN`) will produce an output
   value `NA` (resp. `NaN`). An infinite value `-Inf` (resp. `Inf`)
   correspond to: an output value `0.0` (resp. `0.0`) for the density
   function, an output value `0.0` (resp. `1.0`) for the distribution
   function and `NaN` (resp. `NaN`) for the quantile function.
  
2. When the first argument `p` of the quantile function contains a
   value `< 0.0` or ` > 1.0`, a warning is cast and the corresponding
   output value is `NaN`. When `p` is `0.0` (resp.  `1.0`), the
   corresponding output value is the lower (resp. upper) end-point of
   the distribution, be it finite or not.
  
3. The *parameter arguments* can be vectors and some elements in these
   can be non-finite without resulting in an error. An input value
   `NA` (resp. `NaN`) will produce an output value `NA` (resp. `NaN`).
   When an input value is `-Inf` or `Inf` the corresponding value will
   usually be `NaN`. Yet it can be a finite value in some cases, as in
   `dnorm(x = 1.3, mean = Inf)` which returns `r dnorm(1.3, mean = Inf)`.
   It does not seem wise to formulate a rule about the value
   corresponding to an infinite value of a parameter.

4. When incorrect values are given in a parameter argument such as a
   negative value for a scale argument, the corresponding output value
   is `NaN`, not `NA`. This makes sense as long as `NA` is understood
   as standing for a missing value while `NaN` results from an impossible 
   of ambiguous math operation.

It is worth noting that incorrect or unanticipated values given for
the first argument or parameter arguments never lead to an error as long as
the are numeric. Yet `dnorm("a")` will produce an error.

When computing the log-likelihood of some EV model it is wise to
assume that some observations can be `NA`. The log-likelihood function
of such a model is usually the sum of terms involving the EV density
and possibly the corresponding distribution function - in case of
censoring. One must be sure that a `NA` value in the vector or list of
observations will correspond to a contribution zero in the
log-likelihood. Rule 1. above helps at preventing issues because if
`NA` observations are not omitted as they should, the log-likelihood
will be `NA`.

Rule 2. above matters for EV distributions because it allows the use
of the quantile function to find the end-points which depends on the
parameters. By evaluating the distribution function at these
end-points, one finds the anticipated results.

# Experimentation with Extreme-Value packages

For a number of CRAN Extreme-Value packages, we evaluate the results
given by the probability functions for their two-parameter GP and for
their three-parameter GEV distributions (if any) using non-finite
values in the first argument and possibly for a chosen parameter. The
results are displayed below in tables. Each table corresponds to *an
EV package*, *a distribution* (GPD2 or GEV), *a probability function*
(`d`, `p` or `q`) and *a parameter* (`scale`, `shape`). For each
table, the rows correspond to a value for the chosen parameter and the
columns for the values of the first argument. Mind that the parameter
value for a row can contain a correct value, an incorrect one (e.g., a
negative shape) and a non-finite value. Only one parameter is changed
in each row, the other parameters being kept at their default
value. If no such default value is defined (which turns out to be the
case only for **texmex**) the other parameters are silently given
their "standard" default value as described above.

Note that each row in a table corresponds to a call that is
*vectorized w.r.t. the first argument* (`x` for a density, `q` for a
distribution and `p` for a quantile). So an error must be considered
as holding for a full row. 

## Findings

Our findings are as follows.

- For all packages tested the quantile function `q` casts an error
  when the vector given as probability argument `p` contains
  non-finite values. This behaviour differs from the classical
  behaviour of the quantile function.

- In the packages **evd**, **`POT`** and **SpatialExtremes**, the GPD
  density casts an error `x` is a vector containing non-finite values
  and the shape is `0.0` (the scale being the default `1.0`), but
  not when the shape is changed. This is a rather strange behaviour.

- Most packages do not cope with non-finite values for the parameters
  as the classical distribution functions do. For some packages, the
  behaviour is not consistent across the two distributions or across
  the density and the distribution functions for a same distribution.

- For several packages a (meaningless) finite value can be returned by
  the probability functions corresponding to a first argument `NA` or
  `NaN`. This can lead to difficult-to-understand errors when
  evaluating log-likelihood functions.
	


```{r init, echo=FALSE} 
library(nieve)
library(data.table)

Pars <- list("GEV" = c("loc", "scale", "shape"),
             "GPD2"= c("scale", "shape"))

## First arguments
ProbFuns <- list("d" = list(name = "density", arg = "x"),
                 "p" = list(name = "distribution", arg = "q"),
                 "q" = list(name = "quantile", arg = "p"))

## First argument values
ArgValues <- list("GEV" = list(d = c(1.4, -Inf, Inf, NA, NaN),
                               p = c(1.5, -Inf, Inf, NA, NaN),
                               q = c(-0.1, 0.0, 0.5, 1.0, 1.1, -Inf, Inf, NA, NaN),
                               loc = c(0.0, - Inf, Inf, NA, NaN)),
                  "GPD2" = list(d = c(1.4, -Inf, Inf, NA, NaN),
                                p = c(1.5, -Inf, Inf, NA, NaN),
                                
                                q = c(-0.1, 0.0, 0.5, 1.0, 1.1, -Inf, Inf, NA, NaN)))
## Describe the values of the parameters to be used.
ParValues <- list("GEV" = list(loc = c(0.0, - Inf, Inf, NA, NaN),
                               scale = c(1.0, -1.0, 0.0, - Inf, Inf, NA, NaN),
                               shape = c(-0.4, 0.0, 0.5, - Inf, Inf, NA, NaN)),
                  "GPD2" = list(scale = c(1.0, -1.0, 0.0, - Inf, Inf, NA, NaN),
                                shape = c(-0.4, 0.0, 0.5, - Inf, Inf, NA, NaN)))

## Describe the calls in the different packages, taking into
## account the possibility of different parameter names as in
## `texmex::dgev` and the need of extra parameters as in `extRemes::evd`.
Altern <- list("nieve-GEV" = list(nieve = "GEV",
                                  Suff = "GEV",
                                  Pnm = c(loc = "loc",
                                          scale = "scale",
                                          shape = "shape")),
               "nieve-GPD2" = list(nieve = "GPD2",
                                   Suff = "GPD2",
                                   Pnm = c(scale = "scale",
                                           shape = "shape")),
               "mev-GEV" = list(nieve = "GEV",
                                Suff = "gev",
                                Pnm = c(loc = "loc",
                                        scale = "scale",
                                        shape = "shape")),
               "mev-GPD2" = list(nieve = "GPD2",
                                 Suff = "gp",
                                 Pnm = c(scale = "scale",
                                         shape = "shape")),
               "evd-GEV" = list(nieve = "GEV",
                                Suff = "gev",
                                Pnm = c(loc = "loc",
                                        scale = "scale",
                                        shape = "shape")),
               "evd-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale",
                                        shape = "shape")),
               "POT-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale", shape = "shape")),
               "SpatialExtremes-GPD2" = list(nieve = "GPD2",
                                             Suff = "gpd",
                                             Pnm = c(scale = "scale",
                                                     shape = "shape")),
               "SpatialExtremes-GEV" = list(nieve = "GEV",
                                           Suff = "gev",
                                           Pnm = c(loc = "loc",
                                                   scale = "scale",
                                                   shape = "shape")),
               "tea-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "scale",
                                         shape = "shape")),
               "texmex-GEV" = list(nieve = "GEV",
                                 Suff = "gev",
                                 Pnm = c(loc = "mu",
                                         scale = "sigma",
                                         shape = "xi"),
                                 Def = c(mu = 0.0, sigma = 1.0, xi = 0.0)),
               "texmex-GPD2" = list(nieve = "GPD2",
                                 Suff = "gpd",
                                 Pnm = c(scale = "sigma", shape = "xi"),
                                 Def = c(sigma = 1.0, xi = 0.0)),
               "extRemes-GPD2" = list(nieve = "GPD2",
                                      Suff = "evd",
                                      Pnm = c(scale = "scale", shape = "shape"),
                                      Extra = list(type = "GP")),
               "extRemes-GEV" = list(nieve = "GEV",
                                     Suff = "evd",
                                     Pnm = c(loc = "loc",
                                             scale = "scale",
                                             shape = "shape"),
                                     Extra = list(type = "GEV")),
               "revdbayes-GPD2" = list(nieve = "GPD2",
                                       Suff = "gp",
                                       Pnm = c(scale = "scale", shape = "shape")),
               "revdbayes-GEV" = list(nieve = "GEV",
                                      Suff = "gev",
                                      Pnm = c(loc = "loc",
                                             scale = "scale",
                                             shape = "shape"))
               )

## Put the list in a data frame format using the powerful
## `data.table::rbindlist` function
A <- list()
for (i in seq_along(Altern)) {
    A[[i]] <- data.frame(nieve = Altern[[i]]$nieve,
                         Pkg = gsub("-.*$", "", names(Altern)[i]),
                         Suff = Altern[[i]]$Suff,
                         Pnm = Altern[[i]]$Pnm)
}
AlternDf <- rbindlist(A)
AlternDf <- within(AlternDf, Pkg <- factor(Pkg))

testNonFinite <- function(pkg, print = 0) {

    LMat <- list()
    
    if (print) {
        cat("\n======================\n")
        cat("Package: ", pkg, "\n")
        cat("======================\n")
    }

    dfi <- subset(AlternDf, Pkg == pkg)
    dfi <- dfi[order(dfi$nieve), ]
    
    for (distName in unique(dfi$nieve)) {
        ind <- (dfi$nieve == distName)
        if (print) {
            cat("\no Dist: ", distName, "\n")
        }
        pns <- Pars[[distName]]
        for (pn in pns) {
            if (print) {
                cat("\n  - Parameter:   ", pn, "\n")
            }
            pv <- ParValues[[distName]][[pn]]
            if (print) {
                cat("    Par. values:   ", pv, "\n")
            }
            for (pf in c("d", "p", "q")) {
                ev <- ArgValues[[distName]][[pf]]
                if (print) {
                    cat("\n      Probability function:   ", pf, "\n")
                }
                ## cat("         First argument:   ", format(ev), "\n")
                funAlt <- get(paste0(pf, dfi[ind, ]$Suff[1]),
                              mode = "function",
                              envir = getNamespace(pkg))
                ## cat("         Results:   \n")
                mat <- matrix("", nrow = length(pv), ncol = length(ev))
                rownames(mat) <- paste("       ", pn, "=", format(pv, trim = TRUE))
                colnames(mat) <- paste(ProbFuns[[pf]]$arg, "=", format(ev, trim = TRUE))
                
                for (j in seq_along(pv)) {
                    
                    ## Do we need an ersatz for default values?
                    LI <- Altern[[paste(pkg, distName, sep = "-")]]
                    if (!is.null(LI$Def)) {
                        argAlt <- as.list(LI$Def)
                        argAlt[LI$Pnm[pn]] <- pv[j]
                        add <- list(x = ev)
                        names(add) <- ProbFuns[[pf]]$arg
                        argAlt <- c(add, argAlt)
                    } else {
                        argAlt <- list(ev, pv[j])
                        names(argAlt) <- c(ProbFuns[[pf]]$arg, pn)
                    }
                    
                    if (!is.null(LI$Extra)) {
                       argAlt <- c(argAlt, LI$Extra)
                    }
                    
                    resAlt <- try(do.call(funAlt, argAlt), silent = TRUE)
                    if (!inherits(resAlt, "try-error")) {
                        res <- paste(format(resAlt))
                        mat[j, ] <- format(resAlt)
                    } else {
                        res <- "ERROR"
                        mat[j, ] <- "*****"
                        mat[j, 1] <- "ERROR"
                    }
                }
                nmm <- paste0(pf, distName, "-", pn)
                LMat[[nmm]] <- noquote(mat)
                attr(LMat[[nmm]], "name") <-
                    paste0(pkg, "::", pf, dfi[ind, ]$Suff[1])
            }
        }
    }
    LMat
}
```


```{r, echo=FALSE, results='asis'}
## Many thanks to Yhui Xie for this amazing `knit_expand` function!
src <- lapply(levels(AlternDf$Pkg), function(pkg) {
    knitr::knit_expand('templates/nonFiniteResults.Rmd')
})
res <- knitr::knit_child(text = unlist(src), quiet = TRUE)
cat(res, sep = '\n')
```
